---
title: "Dissecting a WCS Query: DHMV Case Study."
output: rmarkdown::html_vignette
author: "Falk Mielke (falk.mielke@inbo.be)"
vignette: >
  %\VignetteIndexEntry{spatial_dhmv_query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


<a id="org86b1211"></a>

# DHMV Data

Flemish government agencies host a lot of useful web services and data. Some of those are available via web services and can be found here:

-   <https://www.vlaanderen.be/datavindplaats/catalogus>


This vignette demonstrates the **query of elevation data from the Digital Elevation Model**, ``Digitaal HoogteModel Vlaanderen`` (DHMV). We will do it "the complicated way", assembling a web API query. [In the end](#orgd59526e), you will see how to use `inbospatial` to get the same result in a single function call.

This vignette is somewhat related to, but goes beyond the use of Web Feature Services (WFS) for spatial data, [see this tutorial by Thierry Onkelinx](https://tutorials.inbo.be/tutorials/spatial_wfs_services/). In particular, I will document some paths and workarounds to explore if errors occur in the default `ows4R` procedure.

**TL;DR:** see the `?inbospatial:get_coverage_wcs` function to query DHMV (or other WCS) data of a given point in Flanders. An example can be found below.


<a id="org75073fb"></a>

# Situation/Motivation

The purpose of the code below is to query elevation data for an arbitrary location in Flanders. It can be found here:

-   <https://www.vlaanderen.be/datavindplaats/catalogus/wcs-digitaal-hoogtemodel-vlaanderen>

Unfortunately, documentation about the API is sparse, so let's hope it complies with given standards.

We will use Web Coverage Services (WCS), as standardized by the Open Geospatial Consortium (OGC). More info on these services can be found [in the "Geocomputation with R" book, for example](https://r.geocompx.org/read-write#geographic-web-services). The go-to R package for WCS is `ows4R` by Emmanuel Blondel and contributors, [available on github](https://eblondel.github.io/ows4R/).

Unfortunately, neither the geocomputation book's example, nor the vignette in [the `ows4R` package](https://eblondel.github.io/ows4R/articles/wcs.html), nor [the INBO tutorial](https://tutorials.inbo.be/tutorials/spatial_wfs_services/) could be transferred to the DHMV use case. The code simply did not work.

Therefore, I ended up going back to the nitty-gritty aspects of WCS by **manually building a query**.


<a id="orge6d66b9"></a>

# Limits of `ows4R` Package

Before revealing [the solution](#org803f24d), I will document the error and my path to working around it: the core problem was not trivial to see, and you might run into similar issues.

Because, in fact, the `ows4R` package seems functional. In the tutorial mentioned above, [the `ows4R` package](https://CRAN.R-project.org/package=ows4R) is used to query data from web services. However, `ows4R` cannot handle all niche server APIs.

First, we need to load some packages.

```{r setup}
library("httr")
library("sf")
library("terra")
if (!requireNamespace("ows4R", quietly = TRUE)) {
  install.packages("ows4R")
}
library("ows4R")
```

We can connect a WCS client.

```{r}
wcs <- WCSClient$new(
  "https://geo.api.vlaanderen.be/DHMV/wcs", "2.0.1",
  logger = "INFO"
)
```


We can easily query the capabilities:

```{r}
caps <- wcs$getCapabilities()
```

This would grab the same info as you can find in this xml file: <https://geo.api.vlaanderen.be/DHMV/wcs?service=WCS&request=getcapabilities>

Note that the *capabilities* xml holds a lot of valuable information:

-   First and foremost, search it for `CoverageID`; in this case, I was interested in `DHMVII_DTM_1m`: grid coverage of the LIDAR elevation data for Flanders.
-   We can also find that the Coordinate Reference System (CRS) `crsSupported` is EPSG:31370, aka. [Belgian Lambert 72](https://epsg.io/31370). Always good to know.
-   Finally, the capabilities xml provides a `version`: 2.0.1 at the time of writing. Yet keep that in mind for a bit.

The "capabilities" are, per OGC standards, your go-to place for finding API documentation.

In R, we can receive further information conveniently via `ows4R`. This works via the "coverage summary" &#x2026;

```{r}
feature <- "DHMVII_DTM_1m"
dtm_smry <- caps$findCoverageSummaryById(feature, exact = TRUE)
dtm_smry
```

&#x2026; and the coverage description.

```{r}
# Get description from the WCS client
dtm_des <- wcs$describeCoverage(feature)
dtm_des
```


Finally, it is important to know which dimensions are available: `LON/LAT`, `x/y`, or maybe `time`?

```{r}
dtm_dims <- dtm_smry$getDimensions()
dtm_dims
```


So far, so good.

Now, following the tutorials to get some real data:

```{r}
x_test <- 148600
y_test <- 208900
range_m <- 2 # (+/- m)

hopo_boxed <- OWSUtils$toBBOX(
  xmin = x_test - range_m,
  xmax = x_test + range_m,
  ymin = y_test - range_m,
  ymax = y_test + range_m
  )
```

I will write the file to disk:

```{r, eval=FALSE}

mht_file <- tempfile(fileext = ".mht")

tryCatch({
  dtm_data <- dtm_smry$getCoverage(
    bbox = hopo_boxed,
    filename = mht_file
    )
  }, error = function(err) message(err)
)

raster <- rast(mht_file)

## to solve the errors above, see below, or use:
# nolint start
# tif_file <- inbospatial:::unpack_mht(mht_file)
# raster <- rast(tif_file)
# plot(raster, col = gray.colors(256))
# nolint end
```

&#x2026; or, maybe not. The function fails, somehow.

*Status quo:*

-   I receive a file (an "mht" because *I called it* "mht"), supposedly some kind of GeoTIFF image (because I *asked* for `image/tiff`), yet wrapped into other data chunks.
-   I also receive a couple of error messages and warnings:
    -   *"Start tag expected, '<' not found"*
    -   *"cannot open this file as SpatRaster:&#x2026;"*
    -   The raw file is *"not recognized as being in a supported file format. (GDAL error 4)"*

Maybe most useful, `ows4R` briefly displays the URL it used to attempt data query: `https://geo.api.vlaanderen.be/DHMV/wcs?service=WCS&version=2.0.1&coverageId=DHMVII_DTM_1m&subset=x(148599,148601)&subset=y(208899,208901)&format=image/tiff&request=GetCoverage`

Let's break this URL down:

-   the desired `service` is indeed `WCS`,
-   `version` is `2.0.1` (I would guess)
-   `request`, `coverageID`, and `format` seem to be as expected
-   there are `subset` components for the respective dimensions.

Feel free to try yourself to paste the URL to a browser and download the file.

Still, the downloaded file is unreadable. Subtle symptom: even if one adjusts the `subset` (i.e. `bbox` argument, above), the received tif does not change in size; the download is bbox-agnostic, so to speak.

Opening the file with a text editor such as [vim](https://www.vim.org), we see an XML part, some generic binary stuff, some hints that this [should be a GML](https://en.wikipedia.org/wiki/Geography_Markup_Language).

There are [sample geoTIFFs online](https://github.com/mommermi/geotiff_sample/blob/master/sample.tif), and they look different, or maybe not. GML is clearly not geoTIFF, except maybe for the binary part (which I tried to extract, to no avail).

The file does not download with an extension on the author's OS and browser. 
On a different OS, the browser automatically extended it with ".mht".
In fact, it is [an MHT file](https://docs.fileformat.com/web/mht). 
More on that [below](#mht_solution).


<a id="orgfaa5a3d"></a>

# The Lead: QGIS

Well, there is advanced software to open GeoTIFF'esque files. Or geography markup. At least some software might be able to identify what we actually have here.

First, if R failed, Python might work.

```
import rasterio
import rasterio.plot

data_name = "/tmp/test_hopo.tif"
tiff = rasterio.open(data_name)
rasterio.plot.show(tiff, title = "will this work? no!")
```

&#x2026; but it doesn't recognize the file type as raster data.

How about **[QGIS](https://qgis.org)?** *"Spatial Without Compromise"*, they advertise. Unfortunately, that program also was unable to open the file downloaded in R.

However, QGIS can actually connect to WCS directly. This does not help much if you need the data in R. Yet, why not.

Look and behold: QGIS can actually connect to <`https://geo.api.vlaanderen.be/DHMV/wcs`> and query elevation data for Flanders; good indication that the WCS is, somehow, functional. I could zoom and move, export images. I could even export the data. I stopped exporting the data when it threatened to entirely fill my tiny system partition (the raster data is more than 100GB) - good confirmation that we want web services for this.

And, luckily, at some careless map movement, I received an error output from QGIS. With it, QGIS gave me a working URL to query DHMV: 

-  `https://geo.api.vlaanderen.be/DHMV/wcs?SERVICE=WCS&VERSION=1.0.0&REQUEST=GetCoverage&FORMAT=GeoTIFF&COVERAGE=DHMVII_DTM_1m&BBOX=162064,165735,170953,168882&CRS=EPSG:31370&RESPONSE_CRS=EPSG:31370&WIDTH=1622&HEIGHT=574`


```{r echo=FALSE, out.width='90%'}
knitr::include_graphics("../man/figures/qgis_lead.avif")
```

Let's dissect this URL.


<a id="org803f24d"></a>

# The Solution: Query Building

If you start at the question mark of the URL, and split at every ampersand, you can extract the following components from the (correct) QGIS URL:

| **correct:**           |                                   |
|---------------------- |--------------------------------- |
| SERVICE                | WCS                               |
| VERSION                | 1.0.0                             |
| REQUEST                | GetCoverage                       |
| FORMAT                 | GeoTIFF                           |
| COVERAGE               | DHMVII_DTM_1m |
| BBOX                   | 162064,165735,170953,168882       |
| CRS                    | EPSG:31370                        |
| RESPONSE_CRS           | EPSG:31370                        |
| WIDTH                  | 1622                              |
| HEIGHT                 | 574                               |

Compare this to the (unsuccessful) `ows4R` attempt:

| **non-functional:** |                                   |
|------------------- |--------------------------------- |
| service             | WCS                               |
| version             | 2.0.1                             |
| coverageId          | DHMVII_DTM_1m |
| subset              | x(148599,148601)                  |
| subset              | y(208899,208901)                  |
| format              | image/tiff                        |
| request             | GetCoverage                       |

You can selectively shuffle and adjust components, generalizing the working query and learning about the WCS API. Which feels a bit like [fuzzing](https://en.wikipedia.org/wiki/Fuzzing) the `geo.api`.

Here is what I found:

-   `service` (`WCS`) and `request` (`DHMVII_DTM_1m`) were correct; the latter could be swapped for different datasets
-   `coverage` is the correct keyword for reading `DHMVII_DTM_1m`
-   `version` must be `1.0.0`; it currently does not work with 2.0.1
-   instead of subsetting, we can use a `bbox`&#x2026;
-   &#x2026; but a `width` and `height` of the output image are mandatory; alternatively there are `resx` and `resy` (not shown)
-   we must specify a `CRS` (`EPSG:31370`), optionally also a `response_crs`
-   the `format` is `GeoTIFF`; slashes are weird in URLs anyway

After some more trial and error, I could construct working URLs in R and generalize this to a function.

```{r}

get_elevation_wcs <- function(x, y, range_m = 1, tiff_file = NULL) {
  # Please note that this function lacks a lot of
  # assertions and documentation.

  # get bbox
  bbox <- paste(x - range_m, x + range_m,
               y - range_m, y + range_m,
               sep = ",")

  # the URL components
  base_url <- "https://geo.api.vlaanderen.be"
  endpoint <- "/DHMV/wcs" # nolint: absolute_path_linter

  # the query parameters which worked in QGIS
  elevation_query <- list(
    service = "WCS",
    version = "1.0.0",
    request = "GetCoverage",
    format = "GeoTIFF",
    coverage = "DHMVII_DTM_1m",
    bbox = bbox,
    width = 2 * range_m,
    height = 2 * range_m,
    crs = "EPSG:31370"
  )

  # get wcs data
  if (is.null(tiff_file)) {
    tiff_file <- tempfile(fileext = ".tiff")
  }
  GET(url = modify_url(base_url, path = endpoint),
      query = elevation_query,
      write_disk(tiff_file))
  # note: saving this to a file is optional,
  # but might prevent double download or loss of data

  # re-read raster file
  data <- rast(tiff_file)

  return(data)

}
```

Example usage:

```{r}
x_test <- 148600
y_test <- 208900
range_m <- 10 # (+/- m)
test_raster <- get_elevation_wcs(x_test, y_test, range_m = range_m)

# we can extract a value at a point:
extract(test_raster, cbind(x_test, y_test))[[1]]

```


Voila! We can now get the elevation of a given location in Flanders via DHMV web services.

Major grain of salt: we could not use the latest version, `v2.0.1`, of the DHMV API.
This is not a minor limitation, and should be taken seriously: old versions become obsolete, break, or stop data updates.


<a id="mht_solution"></a>

# Using Version `2.0.1`

Apparently, the exact symptoms of the failing query above are somewhat depending on your browser and operating system.
Your browser might detect that the file is an `mht`, or not.


Here is an example of how to build a query with the current version of DHMV:

```{r}

url <- parse_url("https://geo.api.vlaanderen.be/DHMV/wcs")

# the query parameters which worked in QGIS
url$query <- list(
  SERVICE = "WCS",
  VERSION = "2.0.1",
  REQUEST = "GetCoverage",
  FORMAT = "image/tiff",
  COVERAGEID = "DHMVII_DTM_1m",
  SUBSET = "x,http://www.opengis.net/def/crs/EPSG/0/EPSG:31370(148000,149000)",
  SUBSET = "y,http://www.opengis.net/def/crs/EPSG/0/EPSG:31370(208000,209000)",
  SCALEFACTOR = "1",
  CRS = "EPSG:31370",
  RESPONSE_CRS = "EPSG:31370"
)

request <- build_url(url)


# get wcs data
mht_file <- tempfile(fileext = ".mht")

print(request)
print(mht_file)

response <- GET(url = request,
          write_disk(mht_file))
```

With the `mht` specifications [found online](https://docs.fileformat.com/web/mht), it is possible to create an unpacking function, which has become part of this very package.

```{r}
require("readr")
require("stringr")

unpack_mht <- function(mht_filepath) {
  lines_raw <- read_lines_raw(mht_filepath)
  lines_char <- suppressWarnings(read_lines(mht_filepath))
  raw_vector <- read_file_raw(mht_filepath)

  start <- which(str_detect(lines_char, "^II\\*"))
  end <- length(lines_raw) - 1
  pos_start <- length(unlist(lines_raw[1:(start - 1)])) + start
  pos_end <- length(raw_vector) - (length(lines_raw[end + 1]) + 1)

  tif <- raw_vector[pos_start:pos_end]
  tif_path <- str_replace(mht_filepath, "mht", "tif")
  write_file(
    tif,
    tif_path
  )
  return(tif_path)
}
```


This way you get the elevation data extracted:

```{r}
tif_file <- unpack_mht(mht_file)

raster <- rast(tif_file)
```



<a id="orgd59526e"></a>

# The "`inbospatial`" Way

Colleague [Hans Van Calster](https://github.com/hansvancalster) had been facing the same problems with related APIs a year earlier, and he wrote a function for it in [the \`inbospatial\` package](https://github.com/inbo/inbospatial).
We now refined the `get_coverage_wcs()` procedure to solve the `mht` issue.

Here is how to use it, for the same case as above:

```{r}
require("sf")
require("inbospatial")

bbox <- sf::st_bbox(
   c(xmin = 148000, xmax = 149000, ymin = 208000, ymax = 209000),
   crs = sf::st_crs(31370)
)
hopo_raster <- get_coverage_wcs(
  wcs = "DHMV",
  bbox = bbox,
  layername = "DHMVII_DTM_1m",
  version = "2.0.1",
  wcs_crs = "EPSG:31370",
  resolution = 1
)

plot(hopo_raster, col = gray.colors(256))
```

The function gives you quite some options of APIs to query data from.


| **WCS**  | **layer name**           | **reference** |
|------------------- |--------------------------------- |---- |
| dtm  | `EL.GridCoverage.DTM` | [here](https://www.vlaanderen.be/datavindplaats/catalogus/hoogte-dtm)  |
| dsm  | `EL.GridCoverage.DSM` | [here](https://www.vlaanderen.be/datavindplaats/catalogus/hoogte-dsm)  |
| dhmv | `DHMVI_DTM_5m`        | [here](https://metadata.vlaanderen.be/srv/dut/catalog.search#/metadata/9b0f82c7-57c4-463a-8918-432e41a66355)  |
| dhmv | `DHMVII_DTM_1m`       | [here](https://metadata.vlaanderen.be/srv/dut/catalog.search#/metadata/f52b1a13-86bc-4b64-8256-88cc0d1a8735)  |
| dhmv | `DHMVII_DSM_1m`       | [here](https://metadata.vlaanderen.be/srv/dut/catalog.search#/metadata/0da2e5e4-6886-426b-bb82-c0ffe6faeff6)  |

At a quick glance, we found that the `dtm` WCS seems to include canopy points, which is not according to definition; see here: <https://www.neonscience.org/resources/learning-hub/tutorials/chm-dsm-dtm> A general advice is to confirm your data for a region you know.

In addition, there are `oi-omz` and `oi-omw` for "zummer" and winter ortho-mosaic images.

The burden of choice!


<a id="org575ca41"></a>

# *Post Mortem*

The value of this vignette seems to be limited: in essence, it provides the specific query function arguments for elevation model queries with the `geo.api` of the Flemish digital services. Quite a niche use case.

However, this is also a story of tracing errors in existing tools, and scraping other tools for hints.

Can we actually trace the error back to a source? It was pure co-incidence that QGIS provided a working link, or are there general strategies?

In retrospective, the initial issue was worsened by unfavourable circumstances.

-   `ows4R` retrieved a file and produced an error; another error was only thrown downstream. Neither `ows4R`, nor `terra`, `sf`, `rasterio`, or `qgis` were able to handle `mht` files. This lack of support questions the API's design choice of packing and sending an `mht`.
-   Documentation for the `geo.api` is lacking (or I did not find it); in particularly I missed usage examples, vignettes, or tests.

The colleagues at ``Digitaal Vlaanderen`` or the DHMV group certainly had enough work on their table, yet more documentation would be desirable. Same for `ows4R`, judging by their [github issue list](https://github.com/eblondel/ows4R/issues) (see e.g. [this one](https://github.com/eblondel/ows4R/issues/114)). On the other hand, [QGIS is free- and open source](https://en.wikipedia.org/wiki/QGIS), its queries seem to be well maintained and are accessible, errors are transparent, I could have read the source code - a good example piece of software.


Yet in general, if you run into issues like this, do not hesitate to contact support or [file a github issue](https://github.com/eblondel/ows4R/issues/114).
You should not need to fuzz an API.


Thank you for reading, and may all your spatial API queries be successful!


# Table of Contents

- [DHMV Data](#org86b1211)
- [Situation/Motivation](#org75073fb)
- [Limits of `ows4R` Package](#orge6d66b9)
- [The Lead: QGIS](#orgfaa5a3d)
- [The Solution: Query Building](#org803f24d)
- [Using Version `2.0.1`](#mht_solution)
- [The "inbospatial" Way](#orgd59526e)
- [Post Mortem](#org575ca41)

