---
title: "Example use cases to get vector or raster data from web services."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example use cases to get vector or raster data from web services.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(dplyr)
library(inbospatial)
```

# Introduction

Spatial web services allow you to seamlessly pull remote spatial data directly into your R session without manually downloading, unzipping, and managing local files.

Historically, the OGC (Open Geospatial Consortium) established standards like WFS (Web Feature Service) for downloading vector data and WCS (Web Coverage Service) for downloading raster data.
Recently, these older protocols are being actively replaced by the modern, RESTful OGC API framework.
For instance, OGC API Features is the direct, optimized successor to WFS.

In this vignette, we will show how the functions `get_feature_wfs()`, `get_coverage_wcs()`, and the newly added `get_feature_ogc()` can be used to easily fetch data from these services.

For a more in-depth explanation about the underlying mechanics of these services, we refer to the INBO tutorial about  [WFS](https://tutorials.inbo.be/tutorials/spatial_wfs_services/){target="_BLANK"} and the vignette about [WCS](spatial_dhmv_query.html).
The tutorial and vignette take a more hands-on coding approach to manually crafting HTTP requests, whereas in this vignette, much of that complex logic is hidden away safely inside the `inbospatial` functions.

# Overview of Spatial Web Services

Depending on whether you need to actually download the raw data for spatial analysis, or merely view it as a background map, different web services are used.
The table below outlines the most common services, their data types, and how they map to the `inbospatial` package.

| Service Standard | Successor / Modern API | GIS Data Type | Primary Purpose | `inbospatial` Function |
| :--- | :--- | :--- | :--- | :--- |
| **WFS** (Web Feature Service) | OGC API Features | Vector (Points, Lines, Polygons) | **Data Retrieval** | `get_feature_wfs()` |
| **WCS** (Web Coverage Service) | OGC API Coverages | Raster (Gridded/Cell data) | **Data Retrieval** | `get_coverage_wcs()` |
| **OGC API Features** | *Current Standard* | Vector (Points, Lines, Polygons) | **Data Retrieval** | `get_feature_ogc()` |
| **WMS** (Web Map Service) | OGC API Maps | Raster (Rendered Image) | **Visualization** | `add_wms_*()` |
| **WMTS** (Web Map Tile Service) | OGC API Tiles | Vector or Raster Tiles | **Visualization** | `add_wmts_*()` |

# Vector data examples

This section demonstrates both `get_feature_ogc()` and `get_feature_wfs`.
When you have the choice, `get_feature_ogc()` is preferred due to better performance and automatic pagination.
Because spatial servers often limit how much data you can download in a single go, automatic pagination allows the function to detect when more data is available and quietly make follow-up requests until the entire dataset is successfully retrieved for you.

## Simple use case without filter

As a test case, we will query provincial boundaries of Flanders from a layer of the "Provisional Reference of Municipal Boundaries" (`VRBG`, `Voorlopig ReferentieBestand Gemeentegrenzen`), which is available via an API on the `vlaanderen.be` website.

```{r provinces-ogc}
ogc_vrbg <- "https://geo.api.vlaanderen.be/VRBG2025/ogc/features/v1/"

system.time({
  provinces <- get_feature_ogc(
    url = ogc_vrbg,
    collection = "Refprv",
    crs = "EPSG:31370"
  )
})

provinces
```


```{r provinces-wfs, eval=FALSE}
wfs_vrbg <- "https://geo.api.vlaanderen.be/VRBG/wfs"

provinces <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  crs = "EPSG:31370"
)

provinces
```

You can achieve (almost) the same result using a combination of `sf::read_sf()` and `sf::st_transform()`.
For `WFS`, you need to prefix the url with `"WFS:"`.
For `OGC API features`, you need to prefix the url with "`OAPIF`".
We only show the latter.

```{r}
system.time({
  provinces2 <- read_sf(
    paste0("OAPIF:", ogc_vrbg),
    layer = "Refprv"
  )
})

waldo::compare(provinces, provinces2,
  x_arg = "inbospatial", y_arg = "sf",
  max_diffs = 8
)
```

We notice some differences, but nothing substantial.
Going via `sf::read_sf` has added an extra field `id` and renamed `SHAPE` to `geometry`.

Timings do differ substantially.
The `get_feature_ogc()` function should be faster than `sf::read_sf`, because the `OAPIF` method (which is passed on to `GDAL`) relies on the `geoJSON` file format whereas `get_feature_ogc()` collects the data as a `geopackage`.
The latter is a binary file format whereas the former is a text format.
A binary format is much smaller and therefore can be downloaded faster.
It's possible that not all providers of `OGC API features` have implemented `geopackage` as an export option.
If that is the case, you can try `read_sf()` with `OAPIF` instead.

For `WFS`, you may notice other differences as well, most notably for field specifications.
The reason for such differences is that `get_feature_wfs()` retrieves whatever information that is identified from the WFS request - in the same way as you would paste the request in the browser.
This result is then saved temporarily to a `.GML` file and read with `sf::read_sf()`.
Passing the WFS directly to the `dsn` argument of `sf::read_sf`, on the other hand, will translate the request to a form that will pass through [the GDAL WFS driver](https://gdal.org/en/latest/drivers/vector/wfs.html), which handles field specifications by reading them from a `DescribeFeatureType` request:

> At the first opening, the content of the result of the `GetCapabilities` request will be appended to the file, so that it can be cached for later openings of the dataset. The same applies for the `DescribeFeatureType` request issued to discover the field definition of each layer.

## Apply a CQL-Filter to filter on a field

CQL, the "Common Query Language" ([see here, for example](https://docs.geoserver.org/stable/en/user/tutorials/cql/cql_tutorial.html)), provides a way to filter WFS data.
The syntax of [CQL is comparable to SQL](https://www.webcomand.com/docs/language/cql/cql_vs_sql/), yet there are differences in data types and structures.

You can provide and apply filters in `get_feature_ogc()` or `get_feature_wfs()` via the `cql_filter` argument.

```{r west-vlaanderen-ogc}
ogc_vrbg <- "https://geo.api.vlaanderen.be/VRBG2025/ogc/features/v1/"

west_vlaanderen <- get_feature_ogc(
  url = ogc_vrbg,
  collection = "Refprv",
  cql_filter = "NAAM='West-Vlaanderen'",
  crs = "EPSG:31370"
)

west_vlaanderen
```


```{r west-vlaanderen-wfs, eval=FALSE}
wfs_vrbg <- "https://geo.api.vlaanderen.be/VRBG/wfs"

west_vlaanderen_wfs <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  cql_filter = "NAAM='West-Vlaanderen'"
)

west_vlaanderen_wfs
```

Here, we could have also used `sf::read_sf()` in combination with an [OGR SQL query](https://gdal.org/en/latest/user/ogr_sql_dialect.html):

```{r}
west_vlaanderen2 <- read_sf(paste0("OAPIF:", ogc_vrbg),
  query = "SELECT * FROM Refprv WHERE NAAM='West-Vlaanderen'"
)

waldo::compare(west_vlaanderen, west_vlaanderen2)
```

As before, we notice some differences in naming of the fields and the addition of an extra field `id` when using `read_sf()`.

## Request a subset of fields

Instead of querying the complete data table, one can select a subset of variables (fields).
The function `get_feature_ogc()` provides the `properties` argument for it, which accepts a character vector of field names (the field containing the geometry is always returned).
The function `get_feature_wfs()` provides the `property_name` keyword for it, which accepts a string of comma-separated field names.
Note that field names are capitalized on this particular geoserver.

```{r prov-fields-ogc}
ogc_vrbg <- "https://geo.api.vlaanderen.be/VRBG2025/ogc/features/v1/"

prov_fields <- get_feature_ogc(
  url = ogc_vrbg,
  collection = "Refprv",
  crs = "EPSG:31370",
  properties = c("NAAM", "NUTS2")
)

prov_fields
```


```{r prov-fields-wfs, eval=FALSE}
wfs_vrbg <- "https://geo.api.vlaanderen.be/VRBG/wfs"

prov_fields_wfs <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  crs = "EPSG:31370",
  property_name = paste("NAAM", "NUTS2", "SHAPE", sep = ",")
)

prov_fields_wfs
```

Again, we can use `sf::read_sf()` in combination with an [OGR SQL query](https://gdal.org/en/latest/user/ogr_sql_dialect.html):

```{r}
prov_fields_sf <- read_sf(paste0("OAPIF:", ogc_vrbg),
  query = "SELECT NAAM, NUTS2 FROM Refprv"
)

waldo::compare(prov_fields, prov_fields_sf)
```

## Meteorological data example

Another example of a WFS API is the Belgian Royal Meteorological Institute (KMI, `Koninklijk Meteorologisch Instituut van België`).
Details can be [found here](https://opendata.meteo.be/service/wfs?request=GetCapabilities), [and here](https://publish.geo.be/geonetwork/f0ow2say/api/records/RMI_AWS_WFS).

To our knowledge, an `OGC APi features` service is not available for this vector data source.


```{r stations}
kmi <- "https://opendata.meteo.be/service/wfs"

kmi_stations <- get_feature_wfs(
  wfs = kmi,
  layer = "synop:synop_station"
)

kmi_stations
```

This is a list of all KMI stations.
You can plot it with `leaflet` or `mapview` to see the stations on a map and learn that -for instance- Ukkel has code 6447.
So let's extract the maximum daily temperatures for Ukkel starting from the year 2000.

```{r kmi, error=TRUE}
kmi_synop <- get_feature_wfs(
  wfs = kmi,
  layername = "synop:synop_data",
  property_name = paste(
    "code", "timestamp", "temp_max", "the_geom",
    sep = ","
  ),
  cql_filter = "((code = 6447) AND (temp_max IS NOT NULL) AND
    (timestamp >= '2000-01-01 00:00:00.000'))"
)

kmi_synop
```


## Spatial filtering using a bounding box

Few analyses will require a whole countries' data, even if the country is as small as Belgium.
Hence, it might be useful by restricting the data you query with a bounding box.
This works as follows, using a `bbox` with the elements `xmin`, `ymin`, `xmax`, and `ymax`.
For `get_feature_ogc()` you can also provide an object of class `bbox` (recommended) instead of the named vector.
Note: OGC APIs generally expect the `bbox` coordinates to be in WGS84 (EPSG:4326).
If you pass a `bbox` object with another coordinate reference system it will be automatically transformed to WGS84.

```{r bwk-bbox-ogc}
ogc_bwk <- "https://geo.api.vlaanderen.be/BWK/ogc/features/v1"

bbox <- c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900)
bbox <- st_bbox(bbox, crs = st_crs(31370))

bwk_bbox_ogc <- get_feature_ogc(
  url = ogc_bwk,
  collection = "Bwkhab",
  crs = "EPSG:31370",
  bbox = bbox
)

bwk_bbox_ogc

sf::st_geometry(bwk_bbox_ogc) |> plot()
```



```{r bwk-bbox-wfs, eval=FALSE}
wfs_bwk <- "https://geo.api.vlaanderen.be/BWK/wfs"

bwk_bbox_wfs <- get_feature_wfs(
  wfs = wfs_bwk,
  layername = "BWK:Bwkhab",
  crs = "EPSG:31370",
  bbox = c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900)
)

bwk_bbox_wfs

#This returns a `MULTISURFACE` geometry type.
#We can use a couple of `sf` functions to cast this to regular polygons:

bwk_bbox_wfs %>%
  st_cast("GEOMETRYCOLLECTION") %>%
  st_collection_extract("POLYGON") %>%
  st_geometry() %>%
  plot()
```

This can also be done using the `wkt_filter` argument from `sf::read_sf`.

```{r eval=FALSE}
bwk_bbox_sf <- read_sf(
  paste0("OAPIF:", ogc_bwk),
  layer = "Bwkhab",
  wkt_filter = sf::st_as_text(
    sf::st_as_sfc(
      st_bbox(
        c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900),
        crs = sf::st_crs(31370)
      )
    )
  )
)
```


## Spatial filters using spatial binary predicates

In addition to the filters above, you can use logical operations to restrict your area of interest.
For example, features can be selected by intersection with another geographical object.

### Intersects a point

As a proof of concept, we can directly query parameters at a specific location.

```{r intersect-point}
ogc_bwk <- "https://geo.api.vlaanderen.be/BWK/ogc/features/v1"

x_lam <- 173995
y_lam <- 212093

bwk_type <- get_feature_ogc(
  url = ogc_bwk,
  collection = "Bwkhab",
  crs = "EPSG:31370",
  cql_filter = sprintf(
    "S_INTERSECTS(SHAPE,POINT(%s %s))",
    x_lam, y_lam
  ),
  `filter-crs` = "http://www.opengis.net/def/crs/EPSG/0/31370",
)

bwk_type
```

In theory, this should also be possible "the `sf` way".
However, `sf` filters after query (which takes a long time).

```{r eval=FALSE}
bwk_type_sf <- sf::read_sf(
  dsn = paste0("OAPIF:", ogc_bwk),
  layer = "Bwkhab",
  wkt_filter = sprintf("POINT(%s %s)", x_lam, y_lam)
)

bwk_type_sf
```

# Raster data examples

Above, we covered `OGC API features` and Web Feature Services (WFS), which are vector data.
Another common geographical data type are grid-like or "raster" data, which you can retrieve via *Web Coverage Services (WCS)*.
Behold, there is a `get_coverage_wcs()` available for some of the web services we regularly use at INBO.

To know which layers are available in the WCS service, you can use the following code which is based on the `get_wcs_layers()` function.
Here we apply the function to all of the supported WCS services and gather everything in a single table:

```{r}
wcs <- c("dtm", "dsm", "omz", "omw", "dhmv", "mercatornet")

list_of_dfs <- lapply(wcs, function(x) {
  df <- get_wcs_layers(x)
  df <- cbind(wcs = x, df)
  return(df)
})

wcs_layers_df <- do.call(rbind, list_of_dfs)

wcs_layers_df |>
  knitr::kable()
```


Below the example of a false-colour infrared image.
The three bands of the image are visualized using `terra::plotRGB`.

```{r omz-example}
bbox <- sf::st_bbox(
  c(xmin = 99477, xmax = 99541, ymin = 172580, ymax = 172640),
  crs = sf::st_crs(31370)
)
omz <- get_coverage_wcs(
  wcs = "omz",
  bbox = bbox,
  layername = "OI.OrthoimageCoverage.OMZ.CIR",
  resolution = 1,
  version = "1.0.0"
)
terra::plotRGB(omz)
```

Another example, this time extracting the digital terrain model for the same bounding box:

```{r dtm-example}
dtm <- get_coverage_wcs(
  wcs = "dtm",
  bbox = bbox,
  layername = "EL.GridCoverage.DTM",
  resolution = 1
)
terra::plot(dtm)
```

Another example using the `Hillshade` layer from the "dhmv": digital elevation model Flanders:

```{r dhmv-example}
hill <- get_coverage_wcs(
  "dhmv",
  bbox = bbox,
  layername = "DHMV_II_HILL_25cm",
  resolution = 0.25,
  wcs_crs = "EPSG:31370"
)
terra::plot(hill)
```

Finally an example extracting total nitrogen deposition in 2025 at 1 km resolution from the `mercatorNet` service.
Be careful here that you pass the resolution as 1000 meters, i.e. the units of the `wcs_crs`.
We increased the bounding box in this example to a 10 km x 10 km square.

```{r mercatornet-example}
bbox[c("xmax", "ymax")] <- bbox[c("xmax", "ymax")] + 10000

total_nitrogen <- get_coverage_wcs(
  "mercatornet",
  bbox = bbox,
  layername = "hh:hh_vlops_tmd_1km_25MZ23M517",
  resolution = 1000,
  wcs_crs = "EPSG:31370"
)
terra::plot(total_nitrogen)
```


More details on WCS can be found in the `spatial_dhmv_query.Rmd` vignette.
