---
title: "Example use cases to get data from web feature or web coverage services."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wfs_wcs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(dplyr)
library(inbospatial)
```

# Introduction

A WFS or Web Feature Service allows you to download vector spatial data from the web.
A WCS or Web Coverage Service is similar, but is used for raster spatial data.
In this vignette, we will show how to the functions `get_feature_wfs()` and `get_coverage_wcs()` can be used to get data into your R session from these services.

For a more in-depth explanation about WFS services, we refer to [the INBO tutorial about WFS](https://inbo.github.io/tutorials/tutorials/spatial_wfs_services/){target="_BLANK"}.
The tutorial takes a more hands-on coding approach to getting data from a WFS, whereas in this vignette much of the coding stuff is hidden away in the `get_feature_wfs()` function.

# WFS examples

## Simple use case without filter

```{r bel-regions}
wfs_regions <-
  "https://eservices.minfin.fgov.be/arcgis/services/R2C/Regions/MapServer/WFSServer" # nolint

bel_regions <- get_feature_wfs(
  wfs = wfs_regions,
  layername = "regions",
  crs  = "EPSG:31370")

bel_regions

```

You can achieve (almost) the same result using a combination of `sf::read_sf()` and `sf::st_transform()`.

```{r}
bel_regions2 <- read_sf(paste0("WFS:", wfs_regions),
        layer = "regions") %>%
  st_transform(crs = 31370)

waldo::compare(bel_regions, bel_regions2)
```

Most notably, a couple of fields have a different field type and the geometry column is a "MULTIPOLYGON" in the first instance yet a "MULTSURFACE" in the second case.


## Apply a CQL-Filter to filter on a field

```{r prov}
wfs_vrbg <-
  "https://geoservices.informatievlaanderen.be/overdrachtdiensten/VRBG/wfs"

prov <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  cql_filter = "NAAM='West-Vlaanderen'")

prov
```

Here, we could have also used `sf::read_sf()` in combination with an [OGR SQL query](https://gdal.org/user/ogr_sql_dialect.html):

```{r}
prov2 <- read_sf(paste0("WFS:", wfs_vrbg),
                 query = "select * from Refprv where NAAM='West-Vlaanderen'")

waldo::compare(prov, prov2)
```



## Request a subset of fields

```{r prov-fields}
wfs_vrbg <-
  "https://geoservices.informatievlaanderen.be/overdrachtdiensten/VRBG/wfs"

prov_fields <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  crs = "EPSG:31370",
  property_name = paste("NAAM", "NUTS2", "SHAPE", sep = ","))

prov_fields

read_sf(paste0("WFS:", wfs_vrbg),
        query = "select NAAM, NUTS2 from Refprv where NAAM='Antwerpen'")


```

## Spatial filtering using a bounding box

```{r bwk-bbox}
wfs_bwk <-
  "https://geoservices.informatievlaanderen.be/overdrachtdiensten/BWK/wfs"

bwk_bbox <- get_feature_wfs(
  wfs = wfs_bwk,
  layername = "BWK:Bwkhab",
  crs = "EPSG:31370",
  bbox = c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900))


bwk_bbox


bwk_bbox2 <- read_sf(
  paste0("WFS:", wfs_bwk),
  layer = "BWK:Bwkhab",
  wkt_filter = st_as_text(
    st_as_sfc(
      st_bbox(
        c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900),
        crs = st_crs(31370)
      )
    )
  )
)
```

## Spatial filters using spatial binary predicates

### Intersects a point

```{r}
wfs_soiltypes <-
  "https://www.dov.vlaanderen.be/geoserver/bodemkaart/bodemtypes/wfs"

x_lam <- 173995
y_lam <- 212093

soil_type <- get_feature_wfs(
  wfs = wfs_soiltypes,
  layername = "bodemkaart:bodemtypes",
  crs = "EPSG:31370",
  cql_filter = sprintf("INTERSECTS(geom,POINT(%s %s))",
                       x_lam, y_lam))

soil_type


# this is not exactly the same
# the wkt filter takes the bounding box and returns what overlaps with it
read_sf(paste0("WFS:", wfs_soiltypes),
        layer = "bodemkaart:bodemtypes",
        wkt_filter =
          st_point(c(x_lam, y_lam)) %>%
          st_as_text()
          )
```



# WCS examples
