---
title: "Example use cases to get data from web feature or web coverage services."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wfs_wcs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(dplyr)
library(inbospatial)
```

# Introduction

A WFS or Web Feature Service allows you to download vector spatial data from the web.
A WCS or Web Coverage Service is similar, but is used for raster spatial data.
In this vignette, we will show how to the functions `get_feature_wfs()` and `get_coverage_wcs()` can be used to get data into your R session from these services.

For a more in-depth explanation about WFS services, we refer to [the INBO tutorial about WFS](https://inbo.github.io/tutorials/tutorials/spatial_wfs_services/){target="_BLANK"}.
The tutorial takes a more hands-on coding approach to getting data from a WFS, whereas in this vignette much of the coding stuff is hidden away in the `get_feature_wfs()` function.

# WFS examples

## Simple use case without filter

```{r bel-regions}
wfs_regions <-
  "https://eservices.minfin.fgov.be/arcgis/services/R2C/Regions/MapServer/WFSServer" # nolint

bel_regions <- get_feature_wfs(
  wfs = wfs_regions,
  layername = "regions",
  crs = "EPSG:31370"
)

bel_regions
```

You can achieve (almost) the same result using a combination of `sf::read_sf()` and `sf::st_transform()`.

```{r}
bel_regions2 <- read_sf(paste0("WFS:", wfs_regions),
  layer = "regions"
) %>%
  st_transform(crs = 31370)

waldo::compare(bel_regions, bel_regions2,
  x_arg = "inbospatial", y_arg = "sf",
  max_diffs = 8
)
```

There are however a number of differences.
Most notably, a couple of fields have a different field type and the geometry column is a `"MULTIPOLYGON"` in the first instance yet a `"MULTISURFACE"` in the second case.
The reason for these differences is that `get_feature_wfs()` retrieves whatever information that is identified from the WFS request - in the same way as you would paste the request in the browser.
This result is then saved temporarily to a `.GML` file and read with `sf::read_sf()`.
Passing the WFS directly to the `dsn` argument of `sf::read_sf`, on the other hand, will translate the request to a form that will pass through [the GDAL WFS driver](https://gdal.org/drivers/vector/wfs.html), which handles field specifications by reading them from a `DescribeFeatureType` request:

> At the first opening, the content of the result of the `GetCapabilities` request will be appended to the file, so that it can be cached for later openings of the dataset. The same applies for the `DescribeFeatureType` request issued to discover the field definition of each layer.

## Apply a CQL-Filter to filter on a field

```{r prov}
wfs_vrbg <-
  "https://geo.api.vlaanderen.be/VRBG/wfs"

prov <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  cql_filter = "NAAM='West-Vlaanderen'"
)

prov
```

Here, we could have also used `sf::read_sf()` in combination with an [OGR SQL query](https://gdal.org/user/ogr_sql_dialect.html):

```{r}
prov2 <- read_sf(paste0("WFS:", wfs_vrbg),
  query = "select * from Refprv where NAAM='West-Vlaanderen'"
)

waldo::compare(prov, prov2)
```

This time, both methods yield the same result.

## Request a subset of fields

```{r prov-fields}
wfs_vrbg <-
  "https://geo.api.vlaanderen.be/VRBG/wfs"

prov_fields <- get_feature_wfs(
  wfs = wfs_vrbg,
  layername = "VRBG:Refprv",
  crs = "EPSG:31370",
  property_name = paste("NAAM", "NUTS2", "SHAPE", sep = ",")
)

prov_fields
```

Again, we can use `sf::read_sf()` in combination with an [OGR SQL query](https://gdal.org/user/ogr_sql_dialect.html):

```{r}
prov_fields_sf <- read_sf(paste0("WFS:", wfs_vrbg),
  query = "select NAAM, NUTS2 from Refprv"
)

waldo::compare(prov_fields, prov_fields_sf)
```

## Meteorological data example

```{r stations}
kmi <- "https://opendata.meteo.be/service/wfs"

kmi_stations <- get_feature_wfs(
  wfs = kmi,
  layer = "synop:synop_station"
)

kmi_stations
```

This is a list of all KMI stations.
You can plot it with `leaflet` or `mapview` to see the stations on a map and learn that -for instance- Ukkel has code 6447.
So let's extract the maximum daily temperatures for Ukkel starting from the year 2000.

```{r kmi}
kmi_synop <- get_feature_wfs(
  wfs = kmi,
  layername = "synop:synop_data",
  property_name = paste("code", "timestamp", "temp_max",
    "the_geom",
    sep = ","
  ),
  cql_filter = "((code = 6447) AND (temp_max IS NOT NULL) AND
  (timestamp >= '2000-01-01 00:00:00.000'))"
)

kmi_synop
```


## Spatial filtering using a bounding box

```{r bwk-bbox}
wfs_bwk <-
  "https://geo.api.vlaanderen.be/BWK/wfs"

bwk_bbox <- get_feature_wfs(
  wfs = wfs_bwk,
  layername = "BWK:Bwkhab",
  crs = "EPSG:31370",
  bbox = c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900)
)

bwk_bbox
```

This returns a `CURVEPOLYGON` geometry type.
We can use a couple of `sf` functions to cast this to regular polygons:

```{r}
bwk_bbox %>%
  st_cast("GEOMETRYCOLLECTION") %>%
  st_collection_extract("LINESTRING") %>%
  st_cast("POLYGON") %>%
  st_geometry() %>%
  plot()
```

This can also be done using the `wkt_filter` argument from `sf::read_sf`.

```{r eval=FALSE}
bwk_bbox_sf <- read_sf(
  paste0("WFS:", wfs_bwk),
  layer = "BWK:Bwkhab",
  wkt_filter = st_as_text(
    st_as_sfc(
      st_bbox(
        c(xmin = 142600, ymin = 153800, xmax = 146000, ymax = 156900),
        crs = st_crs(31370)
      )
    )
  )
)
```


## Spatial filters using spatial binary predicates

### Intersects a point

```{r}
wfs_soiltypes <-
  "https://www.dov.vlaanderen.be/geoserver/bodemkaart/bodemtypes/wfs"

x_lam <- 173995
y_lam <- 212093

soil_type <- get_feature_wfs(
  wfs = wfs_soiltypes,
  layername = "bodemkaart:bodemtypes",
  crs = "EPSG:31370",
  cql_filter = sprintf(
    "INTERSECTS(geom,POINT(%s %s))",
    x_lam, y_lam
  )
)

soil_type
```

```{r eval=FALSE}
soil_type_sf <- read_sf(
  paste0("WFS:", wfs_soiltypes),
  query = "select * from bodemtypes",
  wkt_filter = sprintf(
    "ST_Intersects(geom,POINT(%s %s))",
    x_lam, y_lam
  )
)

soil_type_sf
```

# WCS examples

Example of a false-colour infrared image.
The three bands of the image are visualized using `terra::plotRGB`.

```{r omz-example}
bbox <- sf::st_bbox(
  c(xmin = 99477, xmax = 99541, ymin = 172580, ymax = 172640),
  crs = sf::st_crs(31370)
)
omz <- get_coverage_wcs(
  wcs = "omz",
  bbox = bbox,
  layername = "OI.OrthoimageCoverage.OMZ.CIR",
  resolution = 1,
  version = "1.0.0"
)
terra::plotRGB(omz)
```

Example of extracting the  digital terrain model for the same bounding box:

```{r dtm-example}
dtm <- get_coverage_wcs(
  wcs = "dtm",
  bbox = bbox,
  layername = "EL.GridCoverage.DTM",
  resolution = 1)
terra::plot(dtm)
```


